Numerical Stability of Detonations in White Dwarf Simulations

### 2018-07-13 papers/ignition/paper.tex
Update timestep limiter

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations. The sensitivity
of simulations to the value of these timestep limiters will be discussed later in the
results. The factors $f_{be}$ and $f_{bX}$ can be set at runtime in \castro.

@@

@@

To understand the consequences of the timestep limiting, and more broadly to 
understand the limitations of Strang splitting, we consider the 
basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
Now, consider that during a head-on collision, initial nuclear burning 
will occur at the contact point between the two stars. Because of 
the staggered updates from splitting, the evolution effectively progresses 
as a cycle between burning for $\Delta t$ and getting fresh material 
advected into the contact point by the hydro update for $\Delta t$. 
When the collision begins, $\Delta t$ is controlled by the hydrodynamic 
stability criterion, and may be large enough that it is possible for 
the burning advance in Step 4 to completely burn the freshly advected 
material all the way to NSE. Consequently the evolution is no longer 
a good approximation to smooth burning of the in-falling material but
rather separate discrete burning and hydro steps, and the nature of 
the burning evolution will be quite different. As the timestep becomes
smaller, the evolution becomes smoother and the splitting error decreases.

>>>>

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
Consider the basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
The principal drawback of this approach is that the reactions and the hydrodynamics
can become decoupled from each other. For example, consider the example of a head-on
collision of white dwarfs. Initial nuclear burning will occur at the contact point
between the two stars. Because of the staggered updates from splitting, the evolution
effectively progresses as a cycle between burning for $\Delta t$ and getting fresh
material advected into the contact point by the hydro update for $\Delta t$.
When the collision begins, $\Delta t$ is controlled by the hydrodynamic stability
criterion, and may be large enough that it is possible for the burning advance in
Step 4 to completely burn the freshly advected material all the way to NSE.
Consequently the evolution is no longer a good approximation to smooth burning of
the in-falling material but rather separate discrete burning and hydro steps, and
the nature of the burning evolution will be quite different.

A common solution to this problem presented in the literature has been to limit the
size of the timestep and thereby limit the extent of this decoupling
\citep{raskin:2010,hawley:2012}, which we adopt here and have implemented in \castro.
Defining the nuclear energy injection timescale $\tau_e$, and the species evolution
timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations.
This limiter is complementary to the limiter based on $f_{be}$, and can catch
certain burning regimes that limiting only based on changes in energy would miss.

@@

The size of the splitting error is a function of the advance timestep, and it
decreases as the timestep becomes smaller. When we add AMR in the regions of
nuclear burning, the hydrodynamic timestep automatically becomes smaller due
to the increased resolution, which mitigates the splitting error. In fact, we
have found that when we have sufficient refinement to achieve a converged
simulation, the size of the hydrodynamic timestep is already small enough
that the splitting error is minimal. Consequently this timestep limiter
is only of great importance for low resolution simulations. Additionally,
we found that for the test problem described later, where we are only trying
to locate the point and time of a detonation rather than assess accurate
nucleosynthetic yields, this timestep limiter has no observable effect,
so we do not discuss it further in this paper.

\section{Timestep Retries}
\label{sec:retries}

@@
